\LARGE \bf Artificial Cognition for Social Human-Robot Interaction:\\ An Implementation

### 2016-07-04 paper.tex
Second round of proof-reading 1

We identify the needed individual and collaborative cognitive skills: geometric
reasoning and situation assessment based on perspective-taking and
affordance analysis; acquisition and representation of knowledge models for
multiple agents (humans and robots, with their specificities); situated,
natural and multi-modal dialogue; human-aware task planning; human-robot
joint task achievement. The article discusses each of these abilities,
presents working implementations, and shows how they combine in a coherent
and original deliberative architecture for human-robot interaction.
Supported by experimental results, we eventually show how explicit knowledge
management, both symbolic and geometric, proves to be instrumental to richer
and more natural human-robot interactions by pushing for pervasive,
human-level semantics within the robot's deliberative system.

@@

    \item a physical environment, estimated through the robot's exteroceptive
        sensing capabilities, and complemented by inferences drawn from previous
        observations,

@@

The main contributions concern the architecture of the decisional layer of a robot designed to share 
the space and the task with humans and to act and interact in a way that facilitates human action and decision.

We present a model of cognitive integration for service robots that:

    \item achieves advanced grounding in complex real environments involving one or
        several humans and a robot;

    \item supports a distributed computation of symbolic knowledge for
        situated dialogue, thanks to the combination of perspective taking,
        affordances computation and logical inference;

@@

    \item reuses the same set of affordances and inferences, together with
        explicit contextual reasoning on humans and robot abilities, to generate
        human-robot shared plans;

    \item exposes a principled approach to integrate a set of complex cognitive
        components in an explicit, semantics-oriented and loosely-coupled
        fashion.

@@

Importantly, this architecture is fully implemented and we demonstrate it on
several robotic platforms and in several actual interaction scenarios. It also proved
to be an effective framework for novel contributions about human-robot joint
action~\cite{fiore2016planning,devin2016implemented,milliez2016using}, as well
as for multi-disciplinary studies~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.

\vspace{0.5cm}
The remaining of the article discusses the robotic architecture that we have
built to address social and autonomous human-robot interaction, and how it
relates to other approaches. We organise this discussion in four sections.

The next section introduces the architecture as a whole, as well as the
knowledge model that we have developed for our robots. Then,
Section~\ref{sec:impl} provides further details on each of the main components of the architecture.
Section~\ref{sec:expe} presents two studies that illustrate in a practical way
what can be currently achieved with our robots.  The
Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally summarise our
main contributions and discuss the key challenges that human-robot interaction
brings to Artificial Intelligence.

@@

Articulating multiple independent software modules in one coherent robotic
architecture is not only a technical challenge, but also a design and
architectural challenge. In particular, properly managing the rich semantics of
human-level natural interaction raises a range of issues. Our basic assumption
and guiding principle is that human-level interaction is easier to achieve if
the robot itself relies internally on human-level semantics.  We implement this
principle by extensively relying on explicit knowledge representation and
manipulation: software components communicate with each other using first-order
logic statements organised into ontologies and whose semantics are close to the
ones manipulated by humans.

Figure~\ref{fig|archi} gives an overview of our architecture. An active
knowledge base ({\sc Oro}), conveniently thought as a semantic
blackboard, connects most of the modules: the geometric
reasoning module ({\sc Spark}) produces at relatively high frequency symbolic assertions (like
\stmt{BOOK1 isOn TABLE}) describing the state of the robot environment and its evolution over time. 
These logical statements are stored in the knowledge base, and
queried back, when necessary, by the language processing module ({\sc Dialogs}), the symbolic task
planner (HATP) and the execution controller. The output of the language
processing module and the activities managed by the robot controller are
likewise stored as symbolic statements.

For instance, when processing a sentence like ``give me another book'', the {\sc
Dialogs} module queries the knowledge base: {\tt \footnotesize find(?book type
Book, ?book differentFrom BOOK1)}\footnote{We present a complete
example in section~\ref{moving-london}.}, and write back assertions like
\stmt{HUMAN desires GIVE\_ACTION45, GIVE\_ACTION45 actsOn BOOK2} to {\sc
Oro}. The HATP planner then uses the knowledge base to initialise the
planning domains with similar requests (like {\tt \footnotesize find(BOOK2 isAt
?location)}), and the execution controller typically monitor
conditions (by subscribing to events like: {\tt \footnotesize
onNewMatch(HUMAN desires ?goal)}) and stores what the robot is currently
doing (\stmt{myself currentlyPerforms GIVE\_ACTION45}).

This architecture moves away from classical layered approaches found in
robotics~\cite{Gat1998three, Volpe2001CLARAty, Goldberg2002}. Contrary to
layered architectures, interactions between components at the deliberative level
are here essentially bidirectional and we do not introduce layers of abstraction
amongst these deliberative components\footnote{We do have lower-level modules to
execute actions or manage sensors, but all cognition-related modules live in the
same global deliberative space.}. The natural language processing component
illustrates this structure: instead of being an independent input modality whose
outputs would be unidirectionally fed to ``higher'' decisional components, it
lives in the deliberative space at the same level as other deliberative
components, and uses the knowledge base in a bidirectional manner, to interpret,
disambiguate natural language, and eventually store newly produced
interpretations (natural language processing is discussed in
Section~\ref{sect|com}).

Our architecture however relates to \emph{Beliefs, Desires, Intentions} (BDI)
architectures. As put by Woolridge~\cite{Woolridge1999}, BDI architectures are
primarily focused on \emph{practical reasoning}, \ie the process of deciding,
step by step, which action to perform to reach a goal. The management of the
interaction between knowledge (the beliefs) and task and plan representation and
execution (the desires and the intentions) is central, and aims at selecting at
each step the best sub-goal. It becomes then an intention that the robot commits
to.

As for any cognitive system, this fundamental interaction between knowledge and
actions is central to our approach as well, and typically involves the dialogue
module to acquire \emph{desires} from the other agents, and the planner and the
execution controller to first decide to take into account (or not) an
incoming desire as a \emph{goal}, and then to generate and manage
\emph{intentions} from these goals through the symbolic task planner.

We extend upon BDI architectures by running other background deliberative task,
without them being explicitly triggered by \emph{desires} in the BDI sense.
The main ones include situation assessment, action monitoring and processing of
non-imperative speech (including performative dialogue that can possibly change
the internal state of the robot, but does not lead directly to the creation of
\emph{desires}, like assertion of new facts or question answering).

@@

In our architecture, knowledge manipulation relies on a central server (the {\sc
Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
knowledge as it is produced by each of the other deliberative components (the
clients). It exposes a {\tt json}-based RPC API to query the knowledge
base~\cite{lemaignan2012kbapi}.  We represent knowledge as RDF triples in the
OWL sub-language~\footnote{\url{http://www.w3.org/TR/owl2-overview/}}. Every
time triples are added or removed from the knowledge base, a Description Logics
reasoner ({\sc Pellet}~\cite{sirin2007pellet})
classifies the whole ontology and inserts all possible inferred triples.  The
clients of the {\sc Oro} server are in charge of managing themselves the
knowledge (when to add, when to update, when to retract knowledge): no
meta-semantics are carried over that would let the server manage itself these
dynamics.\footnote{With one exception: so-called \emph{memory profiles} let the
server automatically discard facts (\ie forget about them) after a specific
period of time. We present this feature in section~\ref{memory}.}

@@

At run-time, the knowledge available to the robot comes from three sources. {\it
A priori} knowledge is stored in an ontology (the {\sc OpenRobots} ontology,
discussed hereafter) and loaded at start-up. This static source covers the
\emph{common-sense} knowledge of the robot, and optionally some
scenario-specific knowledge (for instance, about objects that are to be
manipulated). The second part of the knowledge is acquired at run-time from
perception, interaction and planning. The next sections go into details of these
processes. Lastly, the third source of symbolic statements comes from the
inferences produced by the reasoner.

Contrary to similar projects like KnowRob~\cite{Tenorth2009a} that relies on the
concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
needed, we have an \emph{explicit} approach where we greedily compute and assert
symbolic statements (like spatial relations between objects, see
Section~\ref{sect|sit-ass}). For instance, whenever a client queries KnowRob to
know if \stmt{object1 isOn object2}, KnowRob calls a geometric reasoner to
evaluate if this specific spatial relation holds at that point in time. With our
approach, spatial relations are instead computed and asserted {\it a priori}.
This design choice trades scalability for explicit reasoning: at any time, the
full belief state is made explicit (which may be computationally expensive), and
therefore provides the reasoner with the largest possible inference domain. This
is of special importance for an \emph{event-based} architecture like ours (see
Section~\ref{sect|ctrl}), where, by definition, we do not know beforehand what
we are ``looking for'', and we need instead to greedily infer as much as
possible to build a belief state as comprehensive as possible.

@@

The {\sc Oro} server relies on Description Logics (OWL) to represent and
manipulate knowledge. Relying on RDF triples and Description Logics has advantages such as good
understanding of its trades-off, thanks to being widespread in the semantic Web
community, the availability of mature libraries to manipulate the ontology,
interoperability with several major on-line knowledge bases ({\sc
OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc RoboEarth}~\cite{Waibel2011} are
some examples), open-world reasoning, and the formal guarantee of decidability
(it is always possible to classify a Description Logics ontology).

@@

From the point of view of interactive robotics, it supposes the robot ability to
build, store and retrieve separate models of the beliefs of the humans it
interacts with.  Our knowledge base implements such a mechanism: when the robot
detects that a new human has appeared, it initialises a new independent
knowledge model (an ontology) for this human agent. All the ontologies that are
created share the same common-sense knowledge, but rely on the estimation of the
robot of each agent's perspective for their actual instantiation. For example,
the robot may (geometrically) compute that the book is in its own field of view,
but not in the human one (the detail of this computation, called
\emph{perspective taking}, is discussed in the next section). The robot updates
accordingly the two knowledge models it maintains: the \concept{robot} model is
updated with the fact \stmt{book isVisible true}, while the \concept{human}
model is updated with \stmt{book isVisible false}. These two logical statements
are simultaneously asserted, yet contradict when taken together. Since the two
knowledge models are realised as two independent ontologies, the contradiction
does not actually occur and both the models remain logically consistent.

@@

We rely however on this mechanism in certain cases: some modules
like the natural language processor use the {\it short term} memory profile to
mark concepts that are currently manipulated by the
robot as \emph{active concepts}: if a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{human type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory).
Likewise, recently seen or updated geometric entities are flagged as
\concept{ActiveConcept}. We use this feature during dialogue disambiguation for
instance. On the other hand, our perception layer does not make use of this
mechanism: as described in the next section, the environment model of the robot
is continuously updated and the derived symbolic knowledge is therefore
transient: it lasts only as long as the environment remains in the same state.

@@

\begin{figure}
    \centering
    \begin{tabular}{l}
        \emph{Initial human knowledge} \\
        \hline
        \stmt{book\_1 type Book} \\
        \stmt{human\_1 type Human} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}
        \emph{Input}\\

@@

        \concept{human\_1} says:\\
        ``Give me the book'' \\
        ~\\
        ~\\
        ~\\

@@

        \emph{Generated query to ontology} \\
        \hline
        \concept{find(?obj type Book)} \\ 
        \hspace{0.2cm}$\Rightarrow$ \concept{?obj = book\_1} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}

        \emph{Newly created statements}\\
        \hline
        \stmt{human\_1 desires sit\_1} \\
         \stmt{sit\_1 type Give} \\
         \stmt{sit\_1 performedBy myself} \\
         \stmt{sit\_1 actsOnObject book\_01} \\
         \stmt{sit\_1 receivedBy human\_1} \\
    \end{tabular}

@@

The system supports quantification (``give me \{a | the | some | all | any |
$n$\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), and anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also supports knowledge extension by learning new semantic
structures. For instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
after checking for possible contradictions with existing knowledge. {\sc
Dialogs} finally interprets common temporal and spatial adverbs (like
\emph{above} or \emph{tomorrow}) and translates simple expressions of internal
state  into \emph{experiences} (for instance, ``I'm tired'' is processed into
\stmt{human\_1 experiences state\_1, state\_1 hasFeature tired}, see also
Section~\ref{sect:desires}). A full account of the {\sc Dialogs} features and
the corresponding algorithms is available in~\cite{Lemaignan2011a}.

@@

Note that while the system benefits from complementary modalities, they are not
all required. The system can run with the verbal modality alone, at the cost of
a simpler interaction. For example, if the human says ``this'' without the robot
tracking what the human points at, no \stmt{human\_1 pointsAt ...} fact would
possibly be available in the knowledge base, and the robot falls back on the
anaphora resolution step alone.

@@

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{human1 desires
?action, ?action type Give, ?action actsOnObject ?obj, ?obj type Book}, we could
trigger a behaviour when the human expresses (through dialogue, gestures...)
that he wants the robot to give her a book.

@@

Note also that we take advantage of the reasoning capabilities of the system:
for example, the type of the object (\stmt{?obj type Toy}) may not be
explicitly asserted, but inferred by the reasoner based on other assertions.

@@

When playing this kind of game, however, the issue arises that the robot has no
way to select which knowledge about the object is relevant in the interaction
context. For instance, the knowledge base may store facts like \stmt{obj1 type
ActiveConcept} (which internally means that this concept was mentioned in a
discussion in the last few seconds): this information is not a relevant
property of \concept{obj1} when trying to disambiguate concepts with humans.
This distinction between \emph{internal knowledge} (meaningful to
the system only) and \emph{common knowledge} (whose meaning is understood by
all the interactors) has not been properly dealt with in our architecture.

Besides, even knowledge that belongs to the \emph{common knowledge} may not be
appropriate in a given interaction context. For instance, the system may
compute that at a given instant the human is looking at the object: \stmt{human
looksAt obj1}. This property makes sense to both parties, but in the context of
the \emph{spygame}, we would like to mainly use immanent properties, not
volatile like a gaze. More research is required to identify relevant
interaction contexts and knowledge classes attached to them.

>>>>

We identify first the needed individual and collaborative cognitive skills: geometric
reasoning and situation assessment based on perspective-taking and
affordance analysis; acquisition and representation of knowledge models for
multiple agents (humans and robots, with their specificities); situated,
natural and multi-modal dialogue; human-aware task planning; human-robot
joint task achievement. The article discusses each of these abilities,
presents working implementations, and shows how they combine in a coherent
and original deliberative architecture for human-robot interaction.
Supported by experimental results, we eventually show how explicit knowledge
management, both symbolic and geometric, proves to be instrumental to richer
and more natural human-robot interactions by pushing for pervasive,
human-level semantics within the robot's deliberative system.

@@

    \item a physical environment, estimated through the robot's exteroceptive
        sensing capabilities, and augmented by inferences drawn from previous
        observations,

@@

Our main contributions focus on the architecture of the decisional layer of
social robots. Specifically, the deliberative architecture of a robot designed
to share space and tasks with humans, and to act and interact in a way that
supports the human's own actions and decisions. We present hereafter a model of cognitive
integration for service robots that:

\begin{itemize}
    \item exposes a principled approach to integrate a set of complex cognitive
        components in an explicit, semantics-oriented and yet loosely-coupled
        fashion;

    \item achieves multi-modal and interactive symbol
        grounding~\cite{Harnad1990} in complex, real-world environments involving one or
        several humans and a robot;

    \item distributes the computation of symbolic knowledge by combining perspective taking,
        affordances computation, situated dialogue and logical inference;

@@

    \item reuses the same set of affordances and inferences, together with
        explicit contextual reasoning on humans and robot abilities, to generate
        human-robot shared plans.

@@

This architecture is fully implemented and we demonstrate it on several robotic
platforms and in several interaction scenarios. It eventually proves to be an
effective framework for novel contributions about human-robot joint
action~\cite{fiore2016planning,devin2016implemented,milliez2016using}, as well
as for multi-disciplinary
studies~\cite{dautenhahn2006may,koay2007exploratory,Ros2010b,dehais2011physiological,ferreira2015users,GharbiROMAN2015}.

\vspace{0.5cm} The remaining of the article details this robotic architecture.
We organise this discussion in five sections. The next section introduces the
architecture as a whole, as well as the knowledge model that we have developed
for our robots. Section~\ref{sec:impl} discusses each of the cognitive
components of the architecture.  Section~\ref{sec:expe} presents two studies
that illustrate in a practical way what can be currently achieved with our
robots.  The Sections~\ref{sect|discussion} and~\ref{sect|conclusion} finally
summarise our main contributions and restate the key challenges that human-robot
interaction brings to Artificial Intelligence.

@@

Articulating multiple independent software modules in one coherent robotic
architecture is not only a technical challenge, but also a design and
architectural challenge. In particular, properly managing the rich semantics of
natural interactions with humans raises a range of issues. Our basic assumption
and guiding principle is that human-level interaction is easier to achieve if
the robot itself relies internally on human-level semantics.  We implement this
principle by extensively relying on explicit knowledge representation and
manipulation: software components communicate with each other using first-order
logic statements organised into ontologies and whose semantics are close to the
ones manipulated by humans.

Figure~\ref{fig|archi} gives an overview of our architecture. An active
knowledge base ({\sc Oro}), conveniently thought as a semantic
blackboard, connects most of the modules: the geometric
reasoning module ({\sc Spark}) produces at relatively high frequency symbolic
assertions describing the state of the robot environment and its evolution over
time.  These logical statements are stored in the knowledge base, and queried
back, when necessary, by the language processing module ({\sc Dialogs}), the
symbolic task planner (HATP) and the execution controller ({\sc Shary} or {\sc
pyRobots}). The output of the language processing module and the activities
managed by the robot controller are stored back as symbolic statements as well.

For instance, a book laying on a furniture might be picked up by {\sc Spark} and
represented in symbolic terms as \stmt{BOOK1 type Book, BOOK1 isOn TABLE}. These
symbolic statements are stored in the knowledge base {\sc Oro} and made
available to the other cognitive modules. Later, the robot might process a
sentence like ``give me another book''. The {\sc Dialogs} module would then query the
knowledge base: {\tt \footnotesize find(?obj type Book, ?obj differentFrom
BOOK1)},
and write back assertions like \stmt{HUMAN desires GIVE\_ACTION45,
GIVE\_ACTION45 actsOn BOOK2} to {\sc Oro}. This would in turn trigger the
execution controller {\sc Shary} to prepare to act. It would first call the HATP
planner. The planner uses the knowledge base to initialise the planning domain (\eg
{\tt \footnotesize find(BOOK2 isAt ?location)}), and returns a full symbolic
plan to the execution controller. Finally, the controller would execute the plan and monitor
its achievement, both for itself and for the human. We present complete examples
of similar interactions in Section~\ref{sec:expe}.

Our architecture has not been designed to replicate or provide a
plausible model of human cognition, and in this sense, we distinguish ourselves
from research on \emph{cognitive architectures}. Instead, our main design
principle is to nurture the decisional components of the robot with models of human
behaviour and human preferences in order to develop an effective artifical
cognition for a robot that is able to serve and interact seamlessly with
humans. In this sense, it shares its objectives with~\cite{Fong2005,trafton2013act}.

At high level, this architecture relies on the same principles as a number of
known robotic layered architectures~\cite{Alami98,Gat1998three,Volpe2001CLARAty,
Goldberg2002}. However, the main point for us was to refine and study in detail
the internals of the deliberative layer. In our model, we suggest that the interactions
between components within this deliberative level have to be essentially
bidirectional. We also suggest not to introduce any sub-layers of abstraction amongst these deliberative
components\footnote{We must clarify that we do have lower-level modules to execute actions or manage
sensors, but all cognition-related modules live in the same global deliberative
space.}. The natural language processing component illustrates this structure:
instead of being an independent input modality whose outputs would be
unidirectionally fed to ``higher'' decisional components, it lives in the
deliberative space at the same level as other deliberative components, and make
use of the knowledge base in a bidirectional manner, to interpret, disambiguate natural
language, and eventually store newly produced interpretations (natural language
processing is discussed in Section~\ref{sect|com}). Another example is the
intricate relation between high-level symbolic planning and geometric planning
to deal with affordances and human preferences.

Our architecture relates to \emph{Beliefs, Desires, Intentions} (BDI)
architectures. As put by Woolridge~\cite{Woolridge1999}, BDI architectures are
primarily focused on \emph{practical reasoning}, \ie the process of deciding,
step by step, which action to perform to reach a goal. The management of the
interaction between knowledge (the beliefs) and task and plan representation and
execution (the desires and the intentions) is central, and aims at selecting at
each step the best sub-goal. It becomes then an intention that the robot commits
to. As for any cognitive system, this fundamental interaction between knowledge and
actions is central to our approach as well, and typically involves the dialogue
module to acquire \emph{desires} from the other agents, and the planner and the
execution controller to first decide to take into account (or not) an
incoming desire as a \emph{goal}, and then to generate and manage
\emph{intentions} from these goals through the symbolic task planner.

We extend upon BDI architectures by running other background deliberative tasks,
without them being explicitly triggered by \emph{desires} in the BDI sense.
The main ones include situation assessment, action monitoring and processing of
non-imperative speech (including performative dialogue that can possibly change
the internal state of the robot, but does not lead directly to the creation of
\emph{desires}, like assertion of new facts or question answering).

@@

In our architecture, knowledge manipulation relies on a central server (the {\sc
Oro} server~\cite{Lemaignan2010}, Figure~\ref{fig|spark-oro} top) which stores
knowledge as it is produced by each of the other deliberative components (the
clients). It exposes a {\tt json}-based RPC API to query the knowledge
base~\cite{lemaignan2012kbapi}.  We represent knowledge as RDF triples in the
OWL sub-language~\footnote{\url{http://www.w3.org/TR/owl2-overview/}}. Every
time triples are added or removed from the knowledge base, a Description Logics
reasoner ({\sc Pellet}~\cite{sirin2007pellet})
classifies the whole ontology and inserts all possible inferred triples.  The
clients of the {\sc Oro} server are in charge of managing themselves the
knowledge (when to add, when to update, when to retract knowledge) as no
meta-semantics are carried over that let the server manage itself these
dynamics.\footnote{With one exception: so-called \emph{memory profiles} let the
server automatically discard facts (\ie forget about them) after a specific
period of time. We present this feature in section~\ref{memory}.}

@@

At run-time, the knowledge available to the robot comes from three sources. {\it
A priori} knowledge is stored in an ontology (the {\sc OpenRobots} ontology,
discussed hereafter) and is loaded at start-up. This static source implements the
\emph{common-sense} knowledge of the robot, and might optionally include 
scenario-specific knowledge (for instance, about objects that are to be
manipulated). The second part of the knowledge is acquired at run-time from
perception, interaction and planning. The next sections go into details of these
processes. The third source of symbolic statements comes from the
inferences produced by the reasoner.

Contrary to similar projects like KnowRob~\cite{Tenorth2009a} that relies on the
concept of \emph{computables} to lazily evaluate/acquire symbolic facts when
needed, we have an \emph{explicit} approach where we greedily compute and assert
symbolic statements (like spatial relations between objects, see
Section~\ref{sect|sit-ass}). For instance, whenever a client queries KnowRob to
know if \stmt{OBJECT1 isOn OBJECT2}, KnowRob calls a geometric reasoner to
evaluate if this specific spatial relation holds at that point in time. With our
approach, spatial relations are instead computed and asserted {\it a priori} by
a dedicated process that continuously runs in the background.
This design choice trades scalability for explicit reasoning: at any time, the
full belief state is made explicit, and
therefore provides the reasoner with the largest possible inference domain. This
is of special importance for an \emph{event-based} architecture like ours (see
Section~\ref{sect|ctrl}), where an explicit and comprehensive belief state is
required to not miss events.

@@

The {\sc Oro} server relies on Description Logics (OWL) to represent and
manipulate knowledge. Relying on RDF triples and Description Logics has advantages such as good
understanding of its trades-off, thanks to being widespread in the semantic Web
community; the availability of mature libraries to manipulate the ontology;
interoperability with several major on-line knowledge bases ({\sc
OpenCyc}, {\sc WordNet}, {\sc DBPedia} or {\sc RoboEarth}~\cite{Waibel2011} for
examples); open-world reasoning (which enables us to represent that some facts
might be unknown to the robot); and the formal guarantee of decidability
(it is always possible to classify a Description Logics ontology).

@@

From the point of view of interactive robotics, it supposes the robot ability to
build, store and retrieve separate models of the beliefs of the humans it
interacts with.  Our knowledge base implements such a mechanism: when the robot
detects that a new human has appeared, it initialises a new independent
knowledge model (an ontology) for this human agent. All the ontologies that are
created share the same common-sense knowledge, but rely on the estimation of the
robot of each agent's perspective for their actual instantiation. For example,
the robot may (geometrically) compute that the book is in its own field of view,
but not in the human one (the detail of this computation, called
\emph{perspective taking}, is discussed in the next section). The robot updates
accordingly the two knowledge models it maintains: the \concept{robot} model is
updated with the fact \stmt{BOOK isVisible true}, while the \concept{human}
model is updated with \stmt{BOOK isVisible false}. These two logical statements
are simultaneously asserted, yet contradict when taken together. Since the two
knowledge models are realised as two independent ontologies, the contradiction
does not actually occur and both the models remain logically consistent.

@@

We rely however on this mechanism in certain cases: some modules
like the natural language processor use the {\it short term} memory profile to
mark concepts that are currently manipulated by the
robot as \emph{active concepts}: if a human asks the robot ``Give
me all red objects'', the human, the \concept{Give} action, and each red
objects that are found are successively marked as \emph{active concepts} by
inserting statements such as \stmt{HUMAN type ActiveConcept} in the short-term
memory (which can be considered, in this case, to be a working memory).
Likewise, recently seen or updated geometric entities are flagged as
\concept{ActiveConcept}. We use this feature during dialogue disambiguation for
instance. On the other hand, our perception layer does not make use of this
mechanism: as described in the next section, the environment model of the robot
is continuously updated and the derived symbolic knowledge is therefore
transient: it lasts only as long as the environment remains in the same state.

@@

\begin{figure}
    \centering
    \begin{tabular}{l}
        \emph{Initial human knowledge} \\
        \hline
        \stmt{BOOK1 type Book} \\
        \stmt{HUMAN1 type Human} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}
        \emph{Input}\\

@@

        \concept{HUMAN1} says:\\
        ``Give me the book'' \\
        ~\\
        ~\\
        ~\\

@@

        \emph{Generated query to ontology} \\
        \hline
        \concept{find(?obj type Book)} \\ 
        \hspace{0.2cm}$\Rightarrow$ \concept{?obj = BOOK1} \\
        ~\\
        ~\\
        ~\\
    \end{tabular}
    \begin{tabular}{l}

        \emph{Newly created statements}\\
        \hline
        \stmt{HUMAN1 desires SIT1} \\
         \stmt{SIT1 type Give} \\
         \stmt{SIT1 performedBy myself} \\
         \stmt{SIT1 actsOnObject BOOK1} \\
         \stmt{SIT1 receivedBy HUMAN1} \\
    \end{tabular}

@@

The system supports quantification (``give me \{a | the | some | all | any |
$n$\} can''), thematic roles (action-specific predicates that qualify the
actions), interactive disambiguation (the robot asks questions when it needs
more information), and anaphora resolution (``give \emph{it} to me'') based on
dialogue history. It also supports knowledge extension by learning new semantic
structures. For instance, a sentence like ``learn that cats are animals'' is
converted into \stmt{Cat subClassOf Animal} and added to the knowledge base
after checking for possible contradictions with existing knowledge. {\sc
Dialogs} finally interprets common temporal and spatial adverbs (like
\emph{above} or \emph{tomorrow}) and translates simple expressions of internal
state  into \emph{experiences} (for instance, ``I'm tired'' is processed into
\stmt{HUMAN1 experiences STATE1, STATE1 hasFeature tired}, see also
Section~\ref{sect:desires}). A full account of the {\sc Dialogs} features and
the corresponding algorithms is available in~\cite{Lemaignan2011a}.

@@

Note that while the system benefits from complementary modalities, they are not
all required. The system can run with the verbal modality alone, at the cost of
a simpler interaction. For example, if the human says ``this'' without the robot
tracking what the human points at, no \stmt{HUMAN1 pointsAt ...} fact would
possibly be available in the knowledge base, and the robot falls back on the
anaphora resolution step alone.

@@

This allows us to write reactive robot controllers with a high level of
expressiveness: for instance, by subscribing to the event \stmt{HUMAN1 desires
?action, ?action type Give, ?action actsOnObjecT ?obj, ?obj type Book}, we could
trigger a behaviour when the human expresses (through dialogue, gestures...)
that he wants the robot to give her a book.

@@

Note also that we take advantage of the reasoning capabilities of the system:
for example, the type of the object (\stmt{?obj type Book}) may not be
explicitly asserted, but inferred by the reasoner based on other assertions.

@@

When playing this kind of game, however, the issue arises that the robot has no
way to select which knowledge about the object is relevant in the interaction
context. For instance, the knowledge base may store facts like \stmt{OBJ1 type
ActiveConcept} (which internally means that this concept was mentioned in a
discussion in the last few seconds): this information is not a relevant
property of \concept{obj1} when trying to disambiguate concepts with humans.
This distinction between \emph{internal knowledge} (meaningful to
the system only) and \emph{common knowledge} (whose meaning is understood by
all the interactors) has not been properly dealt with in our architecture.

Besides, even knowledge that belongs to the \emph{common knowledge} may not be
appropriate in a given interaction context. For instance, the system may
compute that at a given instant the human is looking at the object: \stmt{HUMAN
looksAt obj1}. This property makes sense to both parties, but in the context of
the \emph{spygame}, we would like to mainly use immanent properties, not
volatile like a gaze. More research is required to identify relevant
interaction contexts and knowledge classes attached to them.
